Supplementary Materials

### 2025-07-05 point-by-point-response.qmd

Thank you for pointing this out and we agree that the reporting of the mortality outcomes is problematic for a feasibility trial. We have decided to report mortality outcomes descriptively in the paper, given comments from other reviewers, but revised the discussion to focus on the missing data and avoid any conclusions about a potential effect of the training.

## Response to Reviewer 2, Dr. Wietske H.W. Ham, University of Applied Sciences:

> Thank you for submitting this relevant paper and the impressive work behind it. It was an honor to review the paper. Here are some suggestions and comments, which I hope will be helpful in further developing the paper.

Thank you for your comments, they were very helpful. We have revised the manuscript in the light of your comments.

@@

> The flow diagram does not clearly show how many participants did not participate or were missed. The design of the study itself seems to have been conducted properly.

Thank you for pointing this out. We have added the number of patients who were lost to follow up in each arm to the flow diagram.

> In the discussion, deviations from the protocol are mentioned and referred to in the appendix, while these data are actually appropriate to discuss (and share) in the article itself, since it is a pilot.

We have revised the manuscript to describe the deviations where they are most relevant, and retained a list in the supplementary materials for completeness.

@@

> I find the number of appendices in the appendix very extensive, which makes it lose its purpose, and I also wonder to what extent it contributes to answering the research question. Parts of the protocol, which are now included in the appendix, should, in my opinion, be incorporated into the main paper.

We agree that the supplementary materials are very extensive. We included these to transparently report the results of the analyses specified in the original protocol, and we have therefore decided to keep them in the manuscript. We have attempted to include parts of the supplementary materials in the main text where appropriate, in response to your earlier comments.

@@

## Response to Reviewer 3, Dr. Sheila Sprague, McMaster University:

> Thank you for the opportunity to review the manuscript titled “Feasibility of a Cluster Randomised Trial on the Effect of Trauma Life Support Training: A Pilot Study”.  This manuscript describes an important trial with the potential to save many lives at hospitals across India.  There are many sections in this manuscript which are unclear.  The overall structure of the manuscript also needs to be improved upon.  Detailed comments are below.

Thank you for your thorough review. We have made major revisions in response to your comments, and we hope that the manuscript is now clearer.

@@

> Introduction
> - Provide clear objectives for the both the pilot phase and the definitive trial phase

We have revised the objective of the pilot study to also include the aim of the full scale trial. The revised objective is "to assess the feasibility of a cluster randomised controlled trial comparing the effect of ATLS^®^ and PTC with standard care on outcomes in adult trauma patients.".

> Methods
> - Describe what information is being collected on standard of care at each hospital as it varies.

We did not collect in-depth data on the standard care at each hospital, and we have added a note in the methods section to clarify this. We have also added this as a limitation of this pilot study, and suggested that the full scale trial should include a more detailed assessment of the standard care at each hospital.

> - Provide additional details on ATLS and PTC training interventions.

We have added more details on the ATLS and PTC training interventions to the methods section.

> - Include details on the cluster selection.

We have revised the methods section to clarify that the seven hospitals were a convenience sample of hospitals in India that fulfilled the inclusion criteria, and with existing connections to the research team.

> - The authors should comment on the sample size for a pilot study.  For example, why did they select seven clusters? They should also comment on the number of residents and patients required to allow for a meaningful analysis of the feasibility outcomes.

We have added the rationale for the number of clusters and the number of units per cluster to the methods section. We did not conduct a formal power calculation, but deemed that the anticipated sample sizes would be sufficient to evaluate the feasibility outcomes.

> - The reason for the one-month lead in is unclear.  It is unclear how this data will be analyzed.

> - Using sealed envelopes for randomization increases the risk of selection bias, tampering, and accidental loss, potentially compromising allocation concealment. Authors should have used the computer-generated randomization system at time of randomization for each site.
> - For blinding, the authors should state that it was not possible to blind those involved.  It is unclear if those analyzing the data were blinded, as this is possible.
> - The follow-up period for each participant is unclear. The methods state that data were collected for four months.  The methods also mention a 3-day mortality outcome.  
> - Success criteria and threshold details for the feasibility outcomes should be clearly outlined in the methods.
> - The outcomes for the definitive trial (clinical outcomes) need to be clearly listed and described.
> - The statistical analysis section is unclear.  It should list the descriptive analyses plans and then the analyses used to assess the feasibility objectives. Feasibility pilot studies are not powered to detect meaningful differences between groups. Rather their goal is to determine whether a definitive trial is practical by assessing feasibility metrics. The clinical outcomes (the outcomes for the definitive trial) could be reported descriptively, but the use of comparative statistics (differences in final values, differences from baseline, confidence intervals, and subgroup analyses) to assess these outcomes is not appropriate during the pilot phase and potentially misleading, especially without a formal sample size calculation being conducted.
> - The statistical analyses section also describes inconsistent times for outcome comparisons. The intervention arms measure change from baseline over one month pre-training vs. three months post-training, while the control arm measures change over a four-month period. For the definitive trial, the timing of assessment should be consistent across all arms to avoid biased comparisons.

### 2025-07-05 manuscript.Rmd
Continue to work on reviewer 3's comments

Systematic reviews call for controlled trials in settings where these programmes are not routinely implemented [@Jayaraman2014; @Kadhum2020], because conducting such effectiveness trials in settings where they are part of the standard of care is not possible. Many settings without routinely implemented trauma life support training are in low- and middle income countries, where trial logistics can be more challenging. We therefore aimed to assess the feasibility of conducting a cluster randomised controlled trial comparing ATLS^®^ and PTC with standard care.

@@

## Protocol Deviations
Protocol deviations are mentioned where relevant in this manuscript, but a list of all deviations is also included as supplementary material for completeness.

## Trial Design
We piloted a three-armed cluster randomised controlled trial [@GerdinWärnberg2022]. There were a standard care arm and two intervention arms, ATLS^®^ and PTC training. We collected data for four months in all three arms, first during a one month observation phase and then during a three month intervention phase (or continued observation in the standard care arm). Originally, we also aimed to estimate probable effect sizes and other measures needed for the sample size calculations of a full-scale trial [@GerdinWärnberg2022], but we revised this aim in the light of current guidance on the conduct and reporting of pilot trials [ADD CITATION].

## Study Setting
We conducted this pilot study in seven tertiary hospitals across metropolitan areas in India, where neither ATLS^®^, PTC, nor any other established trauma life support training program is routinely taught. Details about each cluster is included as Supplementary Table S1.

@@

Standard care varies across hospitals in India, but most surgical and emergency medicine departments in India organise their physicians in units. These units  include both faculty members and residents, who are assigned a specific day of the week when they are posted in the emergency department. In the emergency department, trauma patients are initially assessed by residents who also resuscitate patients, perform interventions and refer patients for imaging or other investigations. Compared with other settings where a trauma team approach is adopted, nurses and other healthcare professionals are only involved to a limited extent during the initial management.

@@

In each intervention arm the residents in one or two units were trained in either ATLS^®^ or PTC. For the purpose of this pilot study, our target was to train a minimum of 75% of residents in each unit. We did not train the units' faculty, because they are typically not directly involved in the initial management of trauma patients. The ATLS^®^ training was conducted in an ATLS^®^ certified training centre in Mumbai, according to the standard ATLS^®^ curriculum [@acsAtls2018]. The PTC training was conducted in New Delhi, according to the standard PTC curriculum [@ptcfPtc2018]. We did not modify or adapt the delivery or content of these programmes during this pilot study.

@@

We trained resident doctors doing their speciality training in surgery or emergency medicine managing trauma patients in the emergency department and who were expected to remain in the participating hospitals for at least one year from the time of the training. Consent was sought from the residents in each of the intervention groups before they underwent the ATLS^®^ or PTC training.

@@

We measured a large number of outcomes to help plan and assess the feasibility of a full scale trial. A list of outcomes is available in Supplementary Table S1. Our main feasibility outcomes were:

- Consent rate of patients and residents. This was equal to the percentage of patients or residents who consented to be included, out of the total number of eligible patients or residents. 
- Lost to follow up rate. This applied only to patients and was equal to the percentage of patients who did not complete 30 day follow up, out of all enrolled patients.
- Pass rate. This applied only to residents in the intervention arms and was equal the percentage of residents who passed the training programme, out of the total number of trained residents.
- Missing data rate. This applied to each outcome and variable and was equal to the percentage of missing values.
- Differences in distributions of observed and extracted data. This applied to each outcome and variable and compared the distributions of data collected by observations versus extracted from hospital records.

@@

Participating units were screened for eligibility once hospitals confirmed their participation. All residents in these units were approached to consent to training if their hospital was randomised to either of the intervention arms. The training was conducted approximately one month after the study started in that hospital.

@@

We did not conduct a formal power calculation for this pilot study, as the purpose was to assess the feasibility of the trial logistics and research methods. <!--The time period was dictated by budget and time constraints.-->

@@

The research officers collected data on demographics, vital signs, management details including imaging and surgery, and details of any injury sustained. All injuries were coded according to the International Classification of Diseases version 10 (ICD-10). Based on these ICD-10 codes, we calculated the Injury Severity Score using the R package `icdpicr` [@icdpicr_2022]. For a convenience sample of patients we also extracted data from medical records, to be able to compare the distribution of this data with the distribution of data collected through direct observations.<!--Supplemental Material 2 contains a full variable list, with definitions.-->

@@

We conducted community consultations to collect inputs from patients, their caregivers, patient groups, and resident doctors to be used in the selection of outcome measures and implementation of the full-scale trial. The results of these consultations are published separately [@david_patient-reported_2024].

@@

We used the `r version$version.string` Statistical Software for all analyses [@R_2024]. We analysed all data, including the feasibility outcomes, using descriptive statistics and did not perform any formal hypothesis tests [@Lee2014]. Quantitative variables are summarised as median and interquartile range. Qualitative variables are presented as absolute numbers and percentages. Additional analyses performed according to the original protocol are described in the Supplementary Materials and their results shown in Supplementary Tables S1-12..

@@

The percentage of patients who consented to follow up was `r p.consent`% and the percentage of residents who consented to training was `r recruitment.rate.residents`%. Among patients, the lost to follow up rate was `r rate.lost.to.follow.up`%. Among residents, the pass rate was `r pass.rate`%. The missing data rate ranged from `r min.missing.data` to `r max.missing.data`%, with details for selected variables shown in Table \@ref(tab:sample-characteristics) and in Supplementary Table S1. The variables with the maximum amount of missing data were in the cost of treatment, reported in Supplementary Tables S1-12.

@@

We show that it is feasible to conduct and collect data for a cluster randomized controlled trial comparing ATLS^®^ with PTC and standard care. Missing data were low for key variables, including the primary outcome and many secondary outcomes. Some variables, especially the cost of treatment (reported in Supplementary Table S1-12) had very high missing data rates and may not be feasible to include in a full-scale trial, or require different data collection methods. The missing data was substantially higher when data was extracted from medical records instead of being directly observed, but the data were similarly distributed, indicating that data collected from medical records is reliable even if it is less complete.

@@

During the course of this pilot we deviated from the protocol in several ways, and provide a detailed list as Supplementary
material S13. Some key limitations of this pilot and therefore lessons to be learned and factored into the design of the full-scale trial include the lower than expected enrolment rates of some centres, centre specific management routines, and difficulties in collecting data on complications and cause of death.

We attempted to minimse the impact of the lower than expected enrolment rates by including a seventh centre, but careful assessments of patient volumes as part of the screening process will be needed for the full-scale trial. We decided to be pragmatic in selecting which residents to train and how to structure the data collection depending on how and by whom patients were initially managed, but this flexibility will need to be built into the full-scale trial protocol. Finally, we found that data on complications and cause of death were hard to identify and therefore the full-scale trial will need to include longer training of research officers if this data is to be collected.

>>>>

Systematic reviews call for controlled trials in settings where these programmes are not routinely implemented [@Jayaraman2014; @Kadhum2020], because conducting such effectiveness trials in settings where they are part of the standard of care is not possible. Many settings without routinely implemented trauma life support training are in low- and middle income countries, where trial logistics can be more challenging.

We therefore conducted a pilot study to assess the feasibility of a cluster randomised controlled trial comparing the effect of ATLS^®^ and PTC with standard care on outcomes in adult trauma patients.

@@

## Protocol Deviations

Protocol deviations are mentioned where relevant in this manuscript, but a list of all deviations is also included as Supplementary Material S1 for completeness.

## Trial Design

We piloted a three-armed cluster randomised controlled trial [@GerdinWärnberg2022]. There were a standard care arm and two intervention arms, ATLS^®^ and PTC training. We collected data for four months in all three arms, first during a one month observation phase and then during a three month intervention phase (or continued observation in the standard care arm). In the published protocol, we also aimed to estimate probable effect sizes and other measures needed for the sample size calculations of a full-scale trial [@GerdinWärnberg2022], but we revised this aim in the light of current guidance on the conduct and reporting of pilot trials [ADD CITATION].

## Study Setting

We conducted this pilot study in seven tertiary hospitals across metropolitan areas in India, where neither ATLS^®^, PTC, nor any other established trauma life support training program is routinely taught. Details about each cluster is included as Supplementary Table S1. Originally, we intented to include six hospitals as clusters, but decided to include a seventh when they expressed interest and as we had the budget to accommodate this. These seven hospitals were a convenience sample of hospitals in India that fulfilled the inclusion criteria, and with existing connections to the research team.

@@

Standard care varies across hospitals in India, but most surgical and emergency medicine departments in India organise their physicians in units. These units include both faculty members and residents, who are assigned a specific day of the week when they are posted in the emergency department. In the emergency department, trauma patients are initially assessed by either surgical or emergency medicine residents who also resuscitate patients, perform interventions and refer patients for imaging or other investigations. Compared with other settings where a trauma team approach is adopted, nurses and other healthcare professionals are only involved to a limited extent during the initial management. We did not collect data on how standard care varied between the participating hospitals.

@@

In each intervention arm the residents in one or two units were trained in either ATLS^®^ or PTC. For the purpose of this pilot study, our target was to train a minimum of 75% of residents in each unit. We did not train the units' faculty, because they are typically not directly involved in the initial management of trauma patients. The ATLS^®^ training was conducted in an ATLS^®^ certified training centre in Mumbai and the PTC training was conducted in New Delhi. Both trainings were conducted according to their respective standard curriculum [@acsAtls2018; @ptcfPtc2018], and we did not modify or adapt the delivery or content of these programmes during this pilot study.

The provider courses of both programmes are 2.5 days, and they teach how to assess, resuscitate and stabilize trauma patients and how to adapt to needs of different patient populations, for example pediatric, pregnant and geriatric patients. The teaching is based on case discussions and skill stations. There are some important differences between the two programs. The ATLS course include more on the inter-hospital transfer of patients and includes a greater emphasis on the trauma team [@acsAtls2018]. The PTC course on the other hand focuses on trauma care in the low resource setting [@ptcfPtc2018].

@@

We trained resident doctors doing their speciality training in surgery or emergency medicine managing trauma patients in the emergency department and who were expected to remain in the participating hospitals for at least one year from the time of the training. Consent was sought from the residents in each of the intervention groups before they underwent the ATLS^®^ or PTC training. In the published protocol, we stated that only surgical residents would be trained. However, in some of the participating hospitals, emergency medicine residents performed the initial resuscitation and management of trauma patients, and we therefore decided to train them instead.

@@

Our feasibility outcomes were:

- Consent rate of patients and residents. This was equal to the percentage of patients or residents who consented to be included, out of the total number of eligible patients or residents. 
- Lost to follow up rate. This applied only to patients and was equal to the percentage of patients who did not complete 30 day follow up, out of all enrolled patients.
- Pass rate. This applied only to residents in the intervention arms and was equal the percentage of residents who passed the training programme, out of the total number of trained residents.
- Missing data rate. This applied to each outcome and variable and was equal to the percentage of missing values.
- Differences in distributions between directly observed data and data extracted from medical records. This applied to all variables that could be reasonably expected to be present in the medical records, but we only extracted this data from a convenience sample of patients to reduce the workload of the project officers collecting data.

We did not specify success criteria for these outcomes.

@@

Participating units were screened for eligibility once hospitals confirmed their participation. All residents in these units were approached to consent to training if their hospital was randomised to either of the intervention arms. The protocol stated that residents would be approached for consent before the hospitals were randomised, but this proved not to be feasible, and we therefore asked residents for consent after the hospitals were randomised but before training. The training was conducted approximately one month after the study started in that hospital. Initially, we planned to use simple random sampling to select the units to be trained in each intervention hospital, but for pragmatic reasons we decided to leave the decision on which units to train to the site principal investigator.

@@

We wanted to include at least two clusters per arm to avoid drawing conclusions based on single centers, and we wanted to train at least two units per intervention cluster to be able to evaluate the logistics of sending residents for training. We did not conduct a formal power calculation for this pilot study, as the purpose was to assess the feasibility of the trial logistics and research methods. <!--The time period was dictated by budget and time constraints.-->

@@

The research officers collected data on demographics, vital signs, management details including imaging and surgery, and details of any injury sustained. All injuries were coded according to the International Classification of Diseases version 10 (ICD-10). Based on these ICD-10 codes, we calculated the Injury Severity Score using the R package `icdpicr` [@icdpicr_2022]. We also collected data on potential outcomes for the full-scale trial, including mortality within 30-days and in-hospital, complications and cause of death. A list of variables is available in Supplementary Table S2.<!--Supplemental Material 2 contains a full variable list, with definitions.-->

@@

We conducted community consultations to collect inputs from patients, their caregivers, patient groups, and resident doctors to be used in the selection of outcome measures and implementation of the full-scale trial. The results of these consultations are published separately [@david_patient-reported_2024]. We initially planned to distribute periodic surveys to residents and follow them up after 30 days after training, but we changed this to interviews after the end of the study period as we thought this would provide richer data (not published).

@@

We used the `r version$version.string` Statistical Software for all analyses [@R_2024]. We analysed all data, including the feasibility outcomes, using descriptive statistics and did not perform any formal hypothesis tests [@Lee2014]. Initially, we planned to analyse feasibility outcomes both on an overall and individual cluster levels, but the sample sizes in individual clusters were too small to generate meaningful results. Quantitative variables are summarised as median and interquartile range. Qualitative variables are presented as absolute numbers and percentages. Additional analyses performed according to the original protocol are described in the Supplementary Material S2 and their results shown in Supplementary Material S3 and Supplementary Tables S1-12.

@@

The percentage of patients who consented to follow up was `r p.consent`% and the percentage of residents who consented to training was `r recruitment.rate.residents`%. Among patients, the lost to follow up rate was `r rate.lost.to.follow.up`%. Among residents, the pass rate was `r pass.rate`%. The missing data rate ranged from `r min.missing.data` to `r max.missing.data`%, with details for selected variables shown in Table \@ref(tab:sample-characteristics) and in Supplementary Table S2. The variables with the maximum amount of missing data were in the cost of treatment, reported in Supplementary Tables S1-12.

@@

We show that it is feasible to conduct and collect data for a cluster randomized controlled trial comparing ATLS^®^ with PTC and standard care. Missing data were low for key variables, including the primary outcome and many secondary outcomes. Some variables, especially the cost of treatment (reported in Supplementary Table S2-13) had very high missing data rates and may not be feasible to include in a full-scale trial, or require different data collection methods. The missing data was substantially higher when data was extracted from medical records instead of being directly observed, but the data were similarly distributed, indicating that data collected from medical records is reliable even if it is less complete.

@@

During the course of this pilot study we deviated from the protocol in several ways. We have reported these deviations were appropriate in the methods section, and provide a detailed list as Supplementary
material S1. The most important deviation was the revision of the aim of this pilot study, because we initially planned to estimate potential effect sizes and other measures to help sample size calculations for a full scale trial, in addition to the feasibility outcomes. However, current guidance advice against using pilot studies to estimate effect sizes, as the usefulness of these estimates is questionable [INSERT REF]. We therefore decided to only report patient outcomes descriptively. Another important deviation from the protocol was the training of emergency medicine residents. We initially planned to train only surgical residents, but trauma management routines varied between participating hospitals, and we decided to adapt to local routines. The full scale study will need to accommodate this variation as part of the protocol.

There are some important limitations of this pilot study and therefore lessons to be learned and factored into the design of the full-scale trial. First, the volumes of patients included at some of the participating hospitals were lower than expected. A careful assessment of patient volumes as part of the screening process will be needed for the full-scale trial. Second, data on complications and cause of death were almost universally missing. To collect data on these variables will require other methods, as these were not explicitly described in the medical records and autopsy reports are not readily available. Third, we did not collect in-depth data on the standard care at each hospital. This data should be collected as part of the screening process for the full-scale trial.

### 2025-07-05 supplementary-materials.Rmd
Continue to work on reviewer 3's comments

# Protocol Deviations

@@

## Reporting

@@

## Number of Participating Centres

We recruited seven centres instead of six and therefore assigned two centres each to the intervention arms and three centres to the control arm.

@@

# Statistical methods for additional analyses according to the original protocol

@@

# Estimated intracluster correlation coefficient

>>>>

# S1 Protocol Deviations

@@

## Aim

@@

## Number of Participating Clusters

We recruited seven hospitals as clusters instead of six, as a seventh hospital expressed interest and we had the budget to accommodate this request. We therefore assigned two centres each to the intervention arms and three centres to the control arm.

@@

# S2 Statistical methods for additional analyses according to the original protocol

@@

# S3 Estimated intracluster correlation coefficient
